Finch: Sparse and Structured Array Programming with Control Flow

### 2024-04-06 sections/background.tex
fix

\begin{figure}[ht]
\footnotesize
\begin{tabular} {|p{0.7\linewidth}|c|} 
    \hline
    $\finchlookup(seek, body)$: The Lookup Looplet represents a randomly accessible region of an iterator. The body of the lookup is understood to have one less dimension than the lookup itself, as we have already ``looked up'' that index in the tensor by the time we reach the body. \texttt{seek(i)} is a function that updates state to the given index.
    &
    \raisebox{-0.85\totalheight}{\includegraphics[scale=0.20]{Looplets-lookup.png}}
    \\
    \hline
    $\finchrun(body)$: The Run Looplet represents a constant region of an iterator. The body of the run is understood to have one less dimension than the lookup itself, as all of the bodies are identical.
        &
    \raisebox{-0.85\totalheight}{\includegraphics[scale=0.20]{Looplets-run.png}}
    \\ 
    \hline
    $\finchphase(c:d, body)$: The Phase Looplet represents a restriction of the range on which a loop should execute, and allows us to succinctly express the ranges on which children of compound looplets are defined.
    &
    \raisebox{-0.85\totalheight}{\includegraphics[scale=0.20]{Looplets-phase.png}}
    \\ 
    \hline
    $\finchswitch(cond, head, tail)$: The Switch Looplet allows us to specialize the body of a looplet based on a condition, evaluated in the embedding context. If the condition is true, we use `head`, otherwise we use `tail`. Switch has a high lowering priority so we can see what's inside of it and lower that appropriately. This also lifts the condition as high as possible into the loop nest. The condition is assumed to evaluate to a boolean.
    &
    \raisebox{-1\totalheight}{\includegraphics[scale=0.20]{Looplets-switch.png}}
    \\ 
    \hline
    $\finchthunk(preamble, body, epilogue)$: The Thunk Looplet allows us to cache certain computations in the state under which the body will execute. This is useful for computing and caching the results of expensive computations.
    &
    \raisebox{-0.8\totalheight}{\includegraphics[scale=0.20]{Looplets-thunk.png}}
    \\ 
    \hline
    $\finchsequence(head, tail)$: The Sequence looplet represents the concatenation of two looplets. Both arguments must be phase looplets, and are assumed to be nonoverlapping, covering, and in order.
    &
    \raisebox{-0.85\totalheight}{\includegraphics[scale=0.20]{Looplets-sequence.png}}
    \\ 
    \hline
    $\finchspike(body, tail)$: The Spike Looplet represents a run followed by a single value. In this paper, Spike will be considered a shorthand for $\finchsequence(\finchphase(i:j-1, \finchrun(body)), \finchphase(j:j, \finchrun(tail)))$.  In the Finch compiler, spikes are handled with special care, since they are an opportunity to align the final run to the end of the root loop extent, without using any special bounds inference.
    &
    \raisebox{-1.15\totalheight}{\includegraphics[scale=0.20]{Looplets-spike.png}}
    \\ 
    \hline
    $\finchstepper([seek], next, body)$: The stepper looplet represents a variable number of looplets, concatenated. Since our looplets may be skipped over due to conditions or various rewrites, the $seek$ function allows us to fast-forward the state to the start of the root loop extent when it comes time to lower the stepper. The $next$ function advances the state to the next iteration of the stepper. 
    &
    \raisebox{-1\totalheight}{\includegraphics[scale=0.20]{Looplets-stepper.png}}
    \\
    \hline
    \end{tabular}
\vspace{-8pt}
\caption{The looplet language, as understood in a correct execution of a Finch program.}\label{fig:looplets}
\end{figure}

@@

\subsection{Looplets}
Finch represents iteration patterns using Looplets, a language that decomposes datastructure iterators hierarchically. 
%
Looplets represent the control-flow structures needed to iterate over any given datastructure, or multiple datastructures simultaneously. 
%
In particular, looplets are good at lifting code to the highest possible loop level and subdividing iteration hierarchically in coordinate space.
%
Because looplets are compiled with progressive lowering, structure-specific mathematical optimizations such as integrals, multiply by zero, etc. can be implemented using simple compiler passes like term rewriting and constant propagation during the intermediate lowering stages.

The looplets are described in Figure~\ref{fig:looplets}. We simplify the presentation to focus on the semantics, rather than precise implementation. Several looplets introduce or modify variables in the scope of the target language. It is assumed that if a looplet introduces a variable, the child looplet will not modify that variable. For more background on Looplets, we recommend the original work \cite{ahrens_looplets_2023}.

>>>>

\begin{figure}[ht]
\footnotesize
\begin{tabular} {|p{0.7\linewidth}|c|} 
    \hline
    $\finchlookup(seek, body)$: The Lookup looplet represents a randomly accessible region of an iterator. The body of the lookup is understood to have one less dimension than the lookup itself, as we have already ``looked up'' that index in the tensor by the time we reach the body. \texttt{seek(i)} is a function that updates state to the given index.
    &
    \raisebox{-0.85\totalheight}{\includegraphics[scale=0.20]{Looplets-lookup.png}}
    \\
    \hline
    $\finchrun(body)$: The Run looplet represents a constant region of an iterator. The body of the run is understood to have one less dimension than the lookup itself, as all of the bodies are identical.
        &
    \raisebox{-0.85\totalheight}{\includegraphics[scale=0.20]{Looplets-run.png}}
    \\ 
    \hline
    $\finchphase(c:d, body)$: The Phase looplet represents a restriction of the range on which a loop should execute, and allows us to succinctly express the ranges on which children of compound looplets are defined.
    &
    \raisebox{-0.85\totalheight}{\includegraphics[scale=0.20]{Looplets-phase.png}}
    \\ 
    \hline
    $\finchswitch(cond, head, tail)$: The Switch looplet allows us to specialize the body of a looplet based on a condition, evaluated in the embedding context. If the condition is true, we use `head`, otherwise we use `tail`. Switch has a high lowering priority so we can see what's inside of it and lower that appropriately. This also lifts the condition as high as possible into the loop nest. The condition is assumed to evaluate to a Boolean.
    &
    \raisebox{-1\totalheight}{\includegraphics[scale=0.20]{Looplets-switch.png}}
    \\ 
    \hline
    $\finchthunk(preamble, body, epilogue)$: The Thunk looplet allows us to cache certain computations in the state under which the body will execute. This is useful for computing and caching the results of expensive computations.
    &
    \raisebox{-0.8\totalheight}{\includegraphics[scale=0.20]{Looplets-thunk.png}}
    \\ 
    \hline
    $\finchsequence(head, tail)$: The Sequence looplet represents the concatenation of two looplets. Both arguments must be phase looplets, and are assumed to be non-overlapping, covering, and in order.
    &
    \raisebox{-0.85\totalheight}{\includegraphics[scale=0.20]{Looplets-sequence.png}}
    \\ 
    \hline
    $\finchspike(body, tail)$: The Spike looplet represents a run followed by a single value. In this paper, Spike will be considered a shorthand for $\finchsequence(\finchphase(i:j-1, \finchrun(body)), \finchphase(j:j, \finchrun(tail)))$.  In the Finch compiler, spikes are handled with special care, since they are an opportunity to align the final run to the end of the root loop extent, without using any special bounds inference.
    &
    \raisebox{-1.15\totalheight}{\includegraphics[scale=0.20]{Looplets-spike.png}}
    \\ 
    \hline
    $\finchstepper([seek], next, body)$: The stepper looplet represents a variable number of looplets, concatenated. Since our looplets may be skipped over due to conditions or various rewrites, the $seek$ function allows us to fast-forward the state to the start of the root loop extent when it comes time to lower the stepper. The $next$ function advances the state to the next iteration of the stepper. 
    &
    \raisebox{-1\totalheight}{\includegraphics[scale=0.20]{Looplets-stepper.png}}
    \\
    \hline
    \end{tabular}
\vspace{-8pt}
\caption{The Looplet language, as understood in a correct execution of a Finch program.}\label{fig:looplets}
\end{figure}

@@

\subsection{Looplets}
Finch represents iteration patterns using Looplets, a language that decomposes datastructure iterators hierarchically. 
%
Looplets represent the control-flow structures needed to iterate over any given datastructure, or multiple datastructures simultaneously. 
%
In particular, Looplets are good at lifting code to the highest possible loop level and subdividing iteration hierarchically in coordinate space.
%
Because Looplets are compiled with progressive lowering, structure-specific mathematical optimizations such as integrals, multiply by zero, etc. can be implemented using simple compiler passes like term rewriting and constant propagation during the intermediate lowering stages.

The Looplets are described in Figure~\ref{fig:looplets}. We simplify the presentation to focus on the semantics, rather than precise implementation. Several looplets introduce or modify variables in the scope of the target language. It is assumed that if a looplet introduces a variable, the child looplet will not modify that variable. For more background on Looplets, we recommend the original work \cite{ahrens_looplets_2023}.

### 2024-04-06 sections/compiler.tex
fix

The subsequent sections will explain how programs that violate each of these
constraints can be rewritten to programs that satisfy them, and thus how we can
support such a wide variety of programs. For example, we can write nonconcordant
programs like  \mintinline{julia}{for i = _; for j = _; s[] += A[i, j] end end}
by adding a loop to randomly access \mintinline{julia}{A} or adjusting the
storage order of \mintinline{julia}{A} by adding a lazy transposition wrapper.

>>>>

The subsequent sections will explain how programs that violate each of these
constraints can be rewritten to programs that satisfy them, and thus how we can
support such a wide variety of programs. For example, we can write non-concordant
programs like  \mintinline{julia}{for i = _; for j = _; s[] += A[i, j] end end}
by adding a loop to randomly access \mintinline{julia}{A} or adjusting the
storage order of \mintinline{julia}{A} by adding a lazy transposition wrapper.

### 2024-04-06 sections/eval.tex
fix

We evaluate Finch on a broad set of applications to showcase it's efficiency,
flexibility, and expressibility. All of our implementations highlight the
benefits of data structure and algorithm co-design.  Our implementation of
sparse-sparse-matrix multiply (SpGEMM) translates classical lessons from sparse
performance engineering into the language of Finch, using temporaries and
randomly-accessible workspace formats to efficiently implement the three main
approaches. Our study of sparse-matrix-dense-vector multiply (SpMV) highlights
the benefits of precise structural specialization. Our studies of image
morphology and graph applications show how Finch's programming model can express more complex
real-world kernels. Finally, we explain
how flexible operators, formats, and indexing expressions in Finch have
supported a flexible implementation of the Python Array API, supporting fused exection.

>>>>

We evaluate Finch on a broad set of applications to showcase it's efficiency,
flexibility, and expressibility. All of our implementations highlight the
benefits of data structure and algorithm co-design.  Our implementation of
sparse-sparse-matrix multiply (SpGEMM) translates classical lessons from sparse
performance engineering into the language of Finch, using temporaries and
randomly-accessible workspace formats to efficiently implement the three main
approaches. Our study of sparse-matrix-dense-vector multiply (SpMV) highlights
the benefits of precise structural specialization. Our studies of image
morphology and graph applications show how Finch's programming model can express more complex
real-world kernels. Finally, we explain
how flexible operators, formats, and indexing expressions in Finch have
supported a flexible implementation of the Python Array API, supporting fused execution.

### 2024-04-06 sections/eval/morphology.tex
fix

We compared against OpenCV.
%
We used four datasets. We randomly selected 100 images from the MNIST \cite{lecun_gradient-based_1998} and Omniglot \cite{lake_human-level_2015} character recognition datasets, as well as a dataset of human line drawings \cite{eitz_how_2012}. 
%
We also hand-selected a subset of mask images (these images were less homogenous, so we listed them in Appendix C) from a digital image processing textbook \cite{gonzalez_digital_2006}. 
%
All images were thresholded, and we also include versions of the images that have been magnified before thresholding, to induce larger constant regions. 
%
IN our erosion task, the sparseRLE format performs the best as it is asymptotically faster and uses less memory, leading to a 19.5X speedup over OpenCV on the sketches dataset, which becomes arbitrarily large as we magnify the images (here shown as 266X). 
%
We believe the 51.6x on MNIST is due to calling overhead in OpenCV). 
%
The bitwise kernels were effective as well, and would be more effective on datasets with less structure. 
%
A strength of Finch is that it supports structured datasets, even over bitwise operations, allowing us to implement the bitwise kernel and then mask it.

>>>>

We compared against OpenCV.
%
We used four datasets. We randomly selected 100 images from the MNIST \cite{lecun_gradient-based_1998} and Omniglot \cite{lake_human-level_2015} character recognition datasets, as well as a dataset of human line drawings \cite{eitz_how_2012}. 
%
We also hand-selected a subset of mask images (these images were less homogenous, so we listed them in Appendix C) from a digital image processing textbook \cite{gonzalez_digital_2006}. 
%
All images were thresholded, and we also include versions of the images that have been magnified before thresholding, to induce larger constant regions. 
%
In our erosion task, the sparseRLE format performs the best as it is asymptotically faster and uses less memory, leading to a 19.5X speedup over OpenCV on the sketches dataset, which becomes arbitrarily large as we magnify the images (here shown as 266X). 
%
We believe the 51.6x on MNIST is due to calling overhead in OpenCV. 
%
The bitwise kernels were effective as well, and would be more effective on datasets with less structure. 
%
A strength of Finch is that it supports structured datasets, even over bitwise operations, allowing us to implement the bitwise kernel and then mask it.

### 2024-04-06 sections/eval/spmv.tex
fix

  \begin{wrapfigure}{r}{0.5\linewidth}
    \scriptsize
    \centering
    \vspace{-14pt}
    \begin{tabular}{|r|l|l|}
        \hline
        \textbf{Format} & \textbf{Style of Matrix}\\
        \hline
        \mintinline[fontsize=\scriptsize]{julia}{Dense(SparseList(Element(0.0)))} & real, sparse \\
        \mintinline[fontsize=\scriptsize]{julia}{Dense(SparseList(Pattern()))} & boolean, sparse \\
        \mintinline[fontsize=\scriptsize]{julia}{Dense(SparseVBL(Element(0.0)))} & real, blocked \\
        \mintinline[fontsize=\scriptsize]{julia}{Dense(SparseBand(Element(0.0)))} & real, banded \\
        %\mintinline{julia}{Dense(SparsePoint(Element(0.0)))} & real-valued matrices with one value per row \\
        %\mintinline{julia}{Dense(SparseRLE(Pattern(0.0)))} & matrix with runs of true or false \\
        \hline 
    \end{tabular}
      \vspace{-8pt}
    \caption{SpMV Tensor Formats}
    \label{spmv_tensor_formats}
    \vspace{-8pt}
  \end{wrapfigure}

  Figure~\ref{fig:spmv_grouped} displays speedup relative to TACO, SuiteSparseGraphBLAS, and Julia’s standard library.  We test using sparse matrices from a large selection of datasets spanning several previous papers: the datasets used by Vuduc et al. to test the OSKI interface \cite{vuduc2005oski}, Ahrens et al. to test a variable block row format partitioning strategy \cite{ahrens_optimal_2021}, and Kjolstad et al. to test the TACO library \cite{kjolstad_tensor_2017}. Additionally, we included the SNAP graph collection to test with boolean matrices. We also created some synthetic matrices containing bands of varying sizes.
  % or blocks of varying sizes as well as a permutation matrix to encapsulate a few additional use cases.
  %The dense vector is randomly generated. 
  %
  We tested using the row-major and column-major Finch programs in Figure \ref{spmv_programs} as well as the symmetric program where applicable; the performance displayed for Finch on each dataset is the fastest among the formats and programs we tested. For TACO, Column-major SpMV consistently performs better than row-major SpMV (an average of 1.36x better) so we use column-major SpMV in TACO as our baseline.

@@

  We found that the SpMV performance was superior for the level format that best paralleled the structure of the tensor.
  %
  Matrices with a clear blocked structure like exdata\_1, TSOPF\_RS\_b678\_c1, and heart3 performed notably well with the SparseVBL format with speedups of 2.16, 1.55, and 1.30 relative to TACO, while the baseline format had minor slowdowns relative to TACO.
  %
  Furthermore, the synthetic banded matrices we constructed performed the best with the SparseBand matrix, in particular with the large\_band and the medium\_band matrices having a speedup of 1.98 and 1.64 relative to TACO, while the baseline format had minor slowdowns relative to TACO.
  %
  The Pattern format performed better than the Element format for representing the leaf values in the matrices when these values were boolean, such as matrices in the SNAP collection which represent graph datasets are boolean. 
  %
  For example, the SparseList-Pattern for email-Eu-core resulted in a speedup of 2.51, while the SparseList-Element format resulted in a slowdown of 1.84 over TACO.
  %
  Every symmetric matrix in the SparseList and SparseList-Pattern formats has better performance when we use a Finch SpMV program that takes advantage of this symmetry.
  %
  Symmetric SpMV with the SparseList level format in Finch results in an average of 1.27x speedup over TACO and symmetric SpMV with the SparseList-Pattern format in Finch results in an average speedup of 1.21x over TACO.
  %
  Notably, there is a 1.91x speedup for the HB/saylr4 matrix over TACO.

>>>>

  \begin{wrapfigure}{r}{0.5\linewidth}
    \scriptsize
    \centering
    \vspace{-14pt}
    \begin{tabular}{|r|l|l|}
        \hline
        \textbf{Format} & \textbf{Style of Matrix}\\
        \hline
        \mintinline[fontsize=\scriptsize]{julia}{Dense(SparseList(Element(0.0)))} & real, sparse \\
        \mintinline[fontsize=\scriptsize]{julia}{Dense(SparseList(Pattern()))} & Boolean, sparse \\
        \mintinline[fontsize=\scriptsize]{julia}{Dense(SparseVBL(Element(0.0)))} & real, blocked \\
        \mintinline[fontsize=\scriptsize]{julia}{Dense(SparseBand(Element(0.0)))} & real, banded \\
        %\mintinline{julia}{Dense(SparsePoint(Element(0.0)))} & real-valued matrices with one value per row \\
        %\mintinline{julia}{Dense(SparseRLE(Pattern(0.0)))} & matrix with runs of true or false \\
        \hline 
    \end{tabular}
      \vspace{-8pt}
    \caption{SpMV Tensor Formats}
    \label{spmv_tensor_formats}
    \vspace{-8pt}
  \end{wrapfigure}

  Figure~\ref{fig:spmv_grouped} displays speedup relative to TACO, SuiteSparseGraphBLAS, and Julia’s standard library.  We test using sparse matrices from a large selection of datasets spanning several previous papers: the datasets used by Vuduc et al. to test the OSKI interface \cite{vuduc2005oski}, Ahrens et al. to test a variable block row format partitioning strategy \cite{ahrens_optimal_2021}, and Kjolstad et al. to test the TACO library \cite{kjolstad_tensor_2017}. Additionally, we included the SNAP graph collection to test with Boolean matrices. We also created some synthetic matrices containing bands of varying sizes.
  % or blocks of varying sizes as well as a permutation matrix to encapsulate a few additional use cases.
  %The dense vector is randomly generated. 
  %
  We tested using the row-major and column-major Finch programs in Figure \ref{spmv_programs} as well as the symmetric program where applicable; the performance displayed for Finch on each dataset is the fastest among the formats and programs we tested. For TACO, Column-major SpMV consistently performs better than row-major SpMV (an average of 1.36x better) so we use column-major SpMV in TACO as our baseline.

@@

  We found that the SpMV performance was superior for the level format that best paralleled the structure of the tensor.
  %
  Matrices with a clear blocked structure like exdata\_1, TSOPF\_RS\_b678\_c1, and heart3 performed notably well with the SparseVBL format with speedups of 2.16, 1.55, and 1.30 relative to TACO, while the baseline format had minor slowdowns relative to TACO.
  %
  Furthermore, the synthetic banded matrices we constructed performed the best with the SparseBand matrix, in particular with the large\_band and the medium\_band matrices having a speedup of 1.98 and 1.64 relative to TACO, while the baseline format had minor slowdowns relative to TACO.
  %
  The Pattern format performed better than the Element format for representing the leaf values in the matrices when these values were Boolean, such as matrices in the SNAP collection which represent graph datasets are Boolean. 
  %
  For example, the SparseList-Pattern for email-Eu-core resulted in a speedup of 2.51, while the SparseList-Element format resulted in a slowdown of 1.84 over TACO.
  %
  Every symmetric matrix in the SparseList and SparseList-Pattern formats has better performance when we use a Finch SpMV program that takes advantage of this symmetry.
  %
  Symmetric SpMV with the SparseList level format in Finch results in an average of 1.27x speedup over TACO and symmetric SpMV with the SparseList-Pattern format in Finch results in an average speedup of 1.21x over TACO.
  %
  Notably, there is a 1.91x speedup for the HB/saylr4 matrix over TACO.

### 2024-04-06 sections/intro.tex
fix

Our world is full of structured data.
%
Sparse arrays (which store only nonzero elements) describe networks, databases, and simulations~\cite{abhyankarpetsc, bell2007lessons, mcauley2013hidden, balay2020petsc}.
%
Run-length encoding describes images and masks, geometry, and
databases (such as a list of transactions with the date field all the same)~\cite{shi2020column,golomb1966run}.
%
Symmetry, bands, padding, and blocks arise due to modeling choices in scientific computing (e.g., higher order FEMs) as well as in intermediate structures in many linear solvers (e.g., GMRES)~\cite{ded, saad2003iterative, o2009scientific}.
%
In the context of machine learning, combinations of sparse and blocked matrices are increasingly under consideration~\cite{dao2022monarch}.
%
Even complex operators can be expressed as structured arrays.
%
For example, a convolution with a filter can be expressed as a matrix multiplication
with the toeplitz matrix of all the circular shifts of the filter~\cite{sze2017efficient}.

@@

%
Prior implementations are incomplete because the abstractions they use are tightly coupled with the specific data structures that they support.
%
For example, TACO merge lattices represent boolean logic over sets of non-zero values on an integer grid~\cite{kjolstad_tensor_2017}.
%
The polyhedral model allows various compilers to represent dense computations on affine regions~\cite{grosser2012polly}.
%
Taichi enriches single static assignment form with a specialized instruction for accessing only a single sparse structure, but it supports more control flow ~\cite{hu_taichi_2019}.
These systems tightly couple their control flow to narrow classes of data structures to avoid the challenges that occur when we intersect complex control flow with structured data. There are two challenges:
%writing efficient code over structured data

@@

\begin{enumerate}
\item More complex array structures than ever before. We are the first to extend level-by-level hierarchical descriptions to capture banded, triangular, run-length-encoded, or sparse datasets, and any combination thereof.
%
We have chosen a set of level formats that completely captures all combinations of relevant structural properties (zeros, repeated values, and/or blocks).
%
Although many systems (TACO, Taichi, SPF, Ebb)~\cite{chou2018format,  hu_taichi_2019, strout2018sparse, bernstein2016ebb} feature a flexible structure description, our level abstraction is more capable and extensible because it uses Looplets \cite{ahrens_looplets_2023} to express the structure of each level. 
%
\item A rich structured array programming language with for-loops and complex control flow constructs at the same level of productivity of dense arrays. 
%
To our knowledge, the Finch programming language is the first to support if-conditions, early breaks, and multiple left hand sides over structured data, as well as complex accesses such as affine indexing or scatter/gather of sparse or structured operands.
%
\item A compiler that specializes programs to data structures automatically, facilitating an expressive language that makes it easier to search the complex space of algorithms and data structures. Finch reliably utilizes four key properties of structure, such as structural zeros, repeated values, clustered non-zeros, and singletons.
%
\item Our compiler is highly extensible, evidenced by the variety of level formats and control flow constructs that we implement in this work.
%
For example, Finch has been extended to support real-valued array indices with continuous arrays \cite{won2024continuous}.
%
\item We evaluate the efficiency, flexibility, and expressibility of our language in several case studies on a wide range of applications, demonstrating speedups over the state of the art in classic operations such as spmv ($2.51\times$) and spgemm ($1.25\times$), to more complex applications such as graph analytics ($3.53\times$, reducing lines of code by $4\times$ over GraphBLAS), image processing ($19.5\times$), and the Python Array API ($28\times$).
%We also demonstrate how Finch can fuse high-level operations to achieve a significant speedup over non-fused kernels. Additionally, as a case study, a high-level array programming language and fusion interface for operations such as map, broadcast, or reduce that can be compiled to efficient code using the previous loop-level abstractions.
%\item A complete set of level formats for expressing data patterns hierarchically in FiberTree-style decompositions. The first such set of formats to efficiently capture banded, triangular, run-length-encoded, or sparse-run-length-encoded datasets. The formats capture many use cases, from random updates to sequential construction.
%\item The Finch array language, mirroring simple for-loops with imperative code blocks and if-conditions. The first array programming language for the above data formats to support multiple outputs, affine indexing, and imperfectly-nested loops.
%\item Tensor lifecycles, a simple constraint on tensor reads and writes that elegantly restricts Finch programs to avoid complex data dependencies, and enables tensor polymorphism by providing implementers with well-defined functions to overload.
%\item Wrapper Tensors which modify existing datastructures and recombine them to support new patterns, such as affine indexing, padding, transposition, and slicing.
%\item Wrapper Levels which modify existing datastructures and enabling complex features such as atomic updates or contiguous versus separate allocation.
%\item We define the first mappings from the existing pydata/sparse array api high-level operations to low level finch notation
%\item <Performance Contributions>

@@

This advances the state-of-the-art over the looplets work \cite{ahrens_looplets_2023}. While Looplets presented a way to merge iterators over multiple single dimensional structures, we go further by combining single-dimensional looplets into multi-dimensional tensors and operating on them in a programming language with fully-featured control flow.

>>>>

Our world is full of structured data.
%
Sparse arrays (which store only nonzero elements) describe networks, databases, and simulations~\cite{abhyankarpetsc, bell2007lessons, mcauley2013hidden, balay2020petsc}.
%
Run-length encoding describes images and masks, geometry, and
databases (such as a list of transactions with the date field all the same)~\cite{shi2020column,golomb1966run}.
%
Symmetry, bands, padding, and blocks arise due to modeling choices in scientific computing (e.g., higher order FEMs) as well as in intermediate structures in many linear solvers (e.g., GMRES)~\cite{ded, saad2003iterative, o2009scientific}.
%
In the context of machine learning, combinations of sparse and blocked matrices are increasingly under consideration~\cite{dao2022monarch}.
%
Even complex operators can be expressed as structured arrays.
%
For example, a convolution with a filter can be expressed as a matrix multiplication
with the Toeplitz matrix of all the circular shifts of the filter~\cite{sze2017efficient}.

@@

%
Prior implementations are incomplete because the abstractions they use are tightly coupled with the specific data structures that they support.
%
For example, TACO merge lattices represent Boolean logic over sets of non-zero values on an integer grid~\cite{kjolstad_tensor_2017}.
%
The polyhedral model allows various compilers to represent dense computations on affine regions~\cite{grosser2012polly}.
%
Taichi enriches single static assignment form with a specialized instruction for accessing only a single sparse structure, but it supports more control flow ~\cite{hu_taichi_2019}.
These systems tightly couple their control flow to narrow classes of data structures to avoid the challenges that occur when we intersect complex control flow with structured data. There are two challenges:
%writing efficient code over structured data

@@

\begin{enumerate}
\item More complex array structures than ever before. We are the first to extend level-by-level hierarchical descriptions to capture banded, triangular, run-length-encoded, or sparse datasets, and any combination thereof.
%
We have chosen a set of level formats that completely captures all combinations of relevant structural properties (zeros, repeated values, and/or blocks).
%
Although many systems (TACO, Taichi, SPF, Ebb)~\cite{chou2018format,  hu_taichi_2019, strout2018sparse, bernstein2016ebb} feature a flexible structure description, our level abstraction is more capable and extensible because it uses Looplets \cite{ahrens_looplets_2023} to express the structure of each level. 
%
\item A rich structured array programming language with for-loops and complex control flow constructs at the same level of productivity of dense arrays. 
%
To our knowledge, the Finch programming language is the first to support if-conditions, early breaks, and multiple left hand sides over structured data, as well as complex accesses such as affine indexing or scatter/gather of sparse or structured operands.
%
\item A compiler that specializes programs to data structures automatically, facilitating an expressive language that makes it easier to search the complex space of algorithms and data structures. Finch reliably utilizes four key properties of structure, such as structural zeros, repeated values, clustered non-zeros, and singletons.
%
\item Our compiler is highly extensible, evidenced by the variety of level formats and control flow constructs that we implement in this work.
%
For example, Finch has been extended to support real-valued array indices with continuous arrays \cite{won2024continuous}.
%
\item We evaluate the efficiency, flexibility, and expressibility of our language in several case studies on a wide range of applications, demonstrating speedups over the state of the art in classic operations such as SpMV ($2.51\times$) and SpGEMM ($1.25\times$), to more complex applications such as graph analytics ($3.53\times$, reducing lines of code by $4\times$ over GraphBLAS), image processing ($19.5\times$), and the Python Array API ($28\times$).
%We also demonstrate how Finch can fuse high-level operations to achieve a significant speedup over non-fused kernels. Additionally, as a case study, a high-level array programming language and fusion interface for operations such as map, broadcast, or reduce that can be compiled to efficient code using the previous loop-level abstractions.
%\item A complete set of level formats for expressing data patterns hierarchically in FiberTree-style decompositions. The first such set of formats to efficiently capture banded, triangular, run-length-encoded, or sparse-run-length-encoded datasets. The formats capture many use cases, from random updates to sequential construction.
%\item The Finch array language, mirroring simple for-loops with imperative code blocks and if-conditions. The first array programming language for the above data formats to support multiple outputs, affine indexing, and imperfectly-nested loops.
%\item Tensor lifecycles, a simple constraint on tensor reads and writes that elegantly restricts Finch programs to avoid complex data dependencies, and enables tensor polymorphism by providing implementers with well-defined functions to overload.
%\item Wrapper Tensors which modify existing datastructures and recombine them to support new patterns, such as affine indexing, padding, transposition, and slicing.
%\item Wrapper Levels which modify existing datastructures and enabling complex features such as atomic updates or contiguous versus separate allocation.
%\item We define the first mappings from the existing pydata/sparse array api high-level operations to low level finch notation
%\item <Performance Contributions>

@@

This advances the state-of-the-art over the Looplets work \cite{ahrens_looplets_2023}. While Looplets presented a way to merge iterators over multiple single dimensional structures, we go further by combining single-dimensional looplets into multi-dimensional tensors and operating on them in a programming language with fully-featured control flow.
